Think deep for reef resilience

### 2025-09-24 paper/paper_draft.qmd
add line to link to metrics table
Incoporate references based on connectivity model comparison study

Rather than attempting to forecast precise outcomes, the exploratory approach aims to
systematically search many possible futures to identify patterns that lead to outcomes of
interest. The approach allows for the identification of robust management strategies that
perform well across a diversity of future conditions rather than attempting to identify a
single most optimal strategy. Concurrently, assessing the wide swathe of scenarios allows
contextual understanding of the conditions in which strategies may fail. Multiple
interacting and compounding sources of uncertainty, and a lack of broad agreement as to
their likelihood, create inherently uncertain outcomes. Here, exploratory modelling was
applied to identify the environmental and reef characteristics, such as thermal stress,
depth, and connectivity, that support favourable outcomes in reef recovery and resilience.

@@

It is a common conclusion in coral reef ecology that a primary facet of reef resilience is
the supply of recruits that a reef receives following disturbance [@hock2017; @mumby2008;
@torres2024]. Our results suggest that incoming connectivity contributes positively to high
cover cluster allocation for reefs around 15m deep. However, our results suggest no direct,
first order, relationship between weighted incoming connectivity and the ability of reefs to
maintain positive carbonate budgets under future conditions. There has been contrasting
evidence [@bode2018; @hock2017] surrounding whether recent generation connectivity modelling
is able to identify reefs that are important for recovery processes, with findings dependent
on the larval connectivity model used. These investigations were performed using prior
generations of larval connectivity models parameterised for 2008-2013 [@hock2017] and
1996-2002 [@bode2018] time periods, without considering the influence of reef depth on
resilience. Using updated 1km^2^ resolution larval connectivity across the whole GBR, with a
limited temporal sample of particle flows, our results suggest that larval connectivity
alone may not be important or reef resilience and recovery, but should be considered in
combination with more influential factors such as depth. Comprehensive comparisons and
validation of multiple connectivity model outputs would improve certainty and allow for the
targetting of reefs by combining abiotic factors such as depth with connectivity.

Our findings on the importance of connectivity may differ from literature due to the highly
uncertain nature of larval connectivity modelling [@vogt-vincent2023], the way larval flows
and juvenile recruitment are implemented in ADRIA - CoralBlox, and the magnitude of early
bleaching events decreasing the breeding capacity of strongly connected reefs. Analysing
daily connectivity data Vogt-Vincent et al. [-@vogt-vincent2023] suggest that time-mean
connectivity may not be suitable for capturing the seasonal - ecological time scale
connectivity between locations due to high daily variability in connectivity data. This is
supported by the connectivity data used in our investigation, which have large relative
standard deviations for connectivity between reefs over the 24 days used
(@sfig-connectivity-rsd). Using alternative summarisation methods that capture
ecological-timescale patterns in coral larval dispersal, such as cumulated implicit
connectivity [@vogt-vincent2023], future investigations may find increased importance of
both depth and reef connectivity for resilience, in line with other investigations
[@hock2017; @torres2024]. Further improvements and research in connectivity modelling are
necessary when incorporating connectivity as a supplementary factor in reef resilience and
intervention-planning analyses.

@@

Uncertainty in connectivity modelling outputs, satellite-derived bathymetry data and future
thermal stress projections are likely to influence the results obtained in this
investigation. While the resolution of connectivity data used in this investigation improves
upon data used in comparable studies [@bozec2025], the time-mean implementation used may not
effectively capture long term patterns [@vogt-vincent2023]. Using alternative connectivity
implementation methods that better capture the ecological-timescale trends in reef
connections rather than the high daily variability, would improve confidence in the
connectivity findings of this investigation [@vogt-vincent2023]. Additionally greater year -
sample sizes for baseline connectivity data as future computation improvements are made,
along with GBR scale validation efforts, will improve confidence in connectivity data for
ecological modelling and management applications. The DHW data used in this investigation
have not been bias corrected for the Great Barrier Reef region and are part of a wider
Australian dataset [@dixon2022] developed using statistical downscaling. Additionally, Lachs
et al. [@lachs2021] suggest the utility of DHW levels for prediction of coral bleaching can
be improved by reducing the threshold for HotSpot Anomaly values from MMM +1 to MMM and
reducing the accumulation window from 12 weeks to 8 weeks. We investigate a range of future
climate responses, thermal projections and ecological parameterisations of ADRIA - CoralBlox
to account for many possible futures, and our results suggest the factors that influence
reef resilience are consistent across these scenarios.

>>>>

Rather than attempting to forecast precise outcomes, the exploratory approach aims to
systematically search many possible futures to identify patterns that lead to outcomes of
interest. The approach allows for the identification of robust management strategies that
perform well across a diversity of future conditions rather than attempting to identify a
single most optimal strategy. Concurrently, assessing the wide swathe of scenarios allows
contextual understanding of the conditions in which strategies may fail. Multiple
interacting and compounding sources of uncertainty, and a lack of broad agreement as to
their likelihood, create inherently uncertain outcomes. Here, exploratory modelling was
applied to identify the environmental and reef characteristics, such as thermal stress,
depth, and connectivity, that support favourable outcomes in reef recovery and resilience.
Metrics used in analyses can be found in @stbl-metrics.

@@

It is a common conclusion in coral reef ecology that a primary facet of reef resilience is
the supply of recruits that a reef receives following disturbance [@hock2017; @mumby2008;
@torres2024]. Our results suggest that incoming connectivity contributes positively to high
cover cluster allocation for reefs around 15m deep. However, our results suggest no direct,
first order, relationship between weighted incoming connectivity and the ability of reefs to
maintain positive carbonate budgets under future conditions. There has been contrasting
evidence [@bode2018; @hock2017] surrounding whether recent generation connectivity modelling
is able to identify reefs that are important for recovery processes, with findings dependent
on the larval connectivity model used. These investigations were performed using prior
generations of larval connectivity models parameterised for 2008-2013 [@hock2017] and
1996-2002 [@bode2018] time periods, without considering the influence of reef depth on
resilience. Using updated 1km^2^ resolution larval connectivity across the whole GBR, with a
limited temporal sample of particle flows, our results suggest that larval connectivity
alone may not be important or reef resilience and recovery, but should be considered in
combination with more influential factors such as depth. Comprehensive comparisons and
validation of multiple connectivity models, as well as use of a model ensemble, would
improve certainty and allow for the targetting of reefs by combining abiotic factors such as
depth with connectivity.

Our findings on the importance of connectivity likely differ from literature due to the
highly uncertain nature of larval connectivity modelling [@choukroun2025;
@vogt-vincent2023], the way larval flows and juvenile recruitment are implemented in ADRIA -
CoralBlox, and the magnitude of early bleaching events decreasing the breeding capacity of
strongly connected reefs. Analysing daily connectivity data Vogt-Vincent et al.
[-@vogt-vincent2023] suggest that time-mean connectivity may not be suitable for capturing
the seasonal - ecological time scale connectivity between locations due to high daily
variability in connectivity data. This is supported by the connectivity data used in our
investigation, which have large relative standard deviations for connectivity between reefs
over the 24 days used (@sfig-connectivity-rsd). Using alternative summarisation methods that
capture ecological-timescale patterns in coral larval dispersal, such as cumulated implicit
connectivity [@vogt-vincent2023], future investigations may find increased importance of
both depth and reef connectivity for resilience, in line with other investigations
[@hock2017; @torres2024]. Choukroun et al [-@choukroun2025] performed a comparison of three
GBR-scale connectivity models, including GBR1, and suggest the use of a model ensemble
approach to estimate connectivity metrics due to the high sensitivity of connectivity to
model specification. Further improvements and research in connectivity modelling are
necessary when incorporating connectivity as a supplementary factor in reef resilience and
intervention-planning analyses.

@@

Uncertainty in connectivity modelling outputs, satellite-derived bathymetry data and future
thermal stress projections are likely to influence the results obtained in this
investigation. While the resolution of connectivity data used in this investigation improves
upon data used in comparable studies [@bozec2025], the time-mean implementation used may not
effectively capture long term patterns [@vogt-vincent2023]. Using alternative connectivity
implementation methods that better capture the ecological-timescale trends in reef
connections rather than the high daily variability, would improve confidence in the
connectivity findings of this investigation [@vogt-vincent2023]. Additionally greater year -
sample sizes for baseline connectivity data as future computation improvements are made,
along with the derivation of connectivity from an ensemble of biophysical models, will
improve confidence in connectivity data for ecological modelling and management applications
[@choukroun2025]. The DHW data used in this investigation have not been bias corrected for
the Great Barrier Reef region and are part of a wider Australian dataset [@dixon2022]
developed using statistical downscaling. Additionally, Lachs et al. [@lachs2021] suggest the
utility of DHW levels for prediction of coral bleaching can be improved by reducing the
threshold for HotSpot Anomaly values from MMM +1 to MMM and reducing the accumulation window
from 12 weeks to 8 weeks. We investigate a range of future climate responses, thermal
projections and ecological parameterisations of ADRIA - CoralBlox to account for many
possible futures, and our results suggest the factors that influence reef resilience are
consistent across these scenarios.
